DynaProg for Scala

### 2013-01-18 docs/report/conclusion.tex
Nits.

% ------------------------------------------------------------------------------------------------
\section{Future work}
We consider several directions and possible extensions for our work. We briefly describe each of them and give an idea of how they could be implemented:\ol
\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, thereby dramatically reducing the memory transfer penalties introduced by the main memory usage. However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (since we first need to find the enclosing matrix block before addressing the element relatively to it).
\item \textbf{Algorithmic analysis:} so far, we considered that the DSL user would write an optimal program. Another direction in which we could improve the parsers is the recurrence analysis, either by removing serial dependencies when possible (see \ref{calc_simplifications}) or reducing the algorithmic complexity by creating intermediate tabulations (see \ref{user_lang}). These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (hereby knowing how to optimize manually the grammar). 
\item \textbf{Serial problems larger than memory:} As discussed in \ref{serial_memory}, this class of problems require a completely different implementation. Since the authors of \cite{swat_mega} are planning to write extensions to their implementation, duplicating the effort might not be worth the price; however, would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform by George Nithin, a PhD student of the LAP (Laboratory of processors architecture), the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project. The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the memory can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
\item \textbf{Multi-dimensional matrices and independent computations:} In the current implementation, all the matrices are encoded such that they are of the same size. Leveraging the yield analysis, we could reduce the dimension of matrices that are of smaller dimension (for tabulations with bounded maximal size). In the problems we have analyzed, no such special case appeared, this is why we do not support this optimization at present.

>>>>

% ------------------------------------------------------------------------------------------------
\section{Future work}
We consider several directions and possible extensions for our work. We briefly describe each of them and give an idea of how they could be implemented:\ol
\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, thereby dramatically reducing the memory transfer penalties introduced by the main memory usage. However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (since we first need to find the enclosing matrix block before addressing the element relatively to it).
\item \textbf{Algorithmic analysis:} so far, we considered that the DSL user would write an optimal program. Another direction in which we could improve the parsers is the recurrence analysis, either by removing serial dependencies when possible (see \ref{calc_simplifications}) or reducing the algorithmic complexity by creating intermediate tabulations (see \ref{user_lang}). These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (thereby knowing how to optimize manually the grammar).
\item \textbf{Serial problems larger than memory:} As discussed in \ref{serial_memory}, this class of problems require a completely different implementation. Since the authors of \cite{swat_mega} are planning to write extensions to their implementation, duplicating the effort might not be worth the price; however, would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform by George Nithin, a PhD student of the LAP (Laboratory of processors architecture), the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project. The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the memory can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
\item \textbf{Multi-dimensional matrices and independent computations:} In the current implementation, all the matrices are encoded such that they are of the same size. Leveraging the yield analysis, we could reduce the dimension of matrices that are of smaller dimension (for tabulations with bounded maximal size). In the problems we have analyzed, no such special case appeared, this is why we do not support this optimization at present.

### 2013-01-18 docs/report/design.tex
Nits.

Since progress is made along the diagonal of the matrices, a naive row/column addressing strategy would result in no coalesced accesses. Hence we address them by diagonal. Let $M_W$ the width of the matrix and $M_H$ its height. Coalesced addressing is most easy in the parallelogram matrix, as we could pretend that the parallelogram is simply a tilted rectangle  (with diagonal elements being contiguous in memory) as follows:
\[(i,j) \to i + (j-i)*M_H\]

Noticing that the triangle matrix shape is just half the parallelogram, we could use the same addressing scheme, but we would use twice more memory than necessary. Hence we need a more involved formula: the size of the triangle embedded in a square of side $k$ (incl. diagonal) is $\frac{k\cdot(k+1)}{2}$; knowing that fact, the index of the element $(i,j)$ of the triangle (starting with the main diagonal when $i=j$), we obtain the following formula (with $M_H=M_W=n$):
\[(i,j) \to M-T+i \text{ with }\left\{\begin{array}{rcll}
	M&=&\dfrac{(n\cdot (n+1)}{2} \vspace{.15cm} & \textit{total }\bigtriangleup\\
	d&=&n+1+i-j \vspace{.15cm} & \textit{diagonal of }(i,j)\\
	T &=&\dfrac{(d\cdot (d+1)}{2} & \bigtriangleup \textit{ of current diagonal}
\end{array}\right.\]

Finally, for the rectangular matrix, since the parallelogram indexing looks efficient, we want to reuse the same idea. However, embedding the rectangle within a parallelogram would have a very large overhead ($(M_H)^2$, by adding a triangle on each side of the rectangle). The solution consists into breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, thereby dramatically reducing the size of the additional triangles. Finally, the stripes can be stitched together to form a single parallelogram continuing along the next stripe. Noticing that beyond a certain number of lines, coalescing access does not improve latency as memory is anyway not stored contiguously, we can make $B_H$ constant .

@@

% ------------------------------------------------------------------------------------------------
\subsection{LMS integration} \label{lms_use}
At first glance, LMS seems the ideal candidate to transform Scala code into its C-like equivalent. However, the concern in this project is that the GPU code sensibly differers from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.

Ad-hoc C translation seems more appropriate than using LMS for the following reasons: \ul
\item LMS uses the type {\tt Rep[T]} to represent a lifted object of type {\tt T}. We tried to use LMS to generate both the Scala and CUDA program. This approach suffered from multiple complication from types, in particular it becomes very tedious to break a lifted tuple into a tuple of lifted values (that is convert {\tt Rep[(A,B,C)]} into {\tt (Rep[A],Rep[B],Rep[C])}). This issue is particularly annoying in our particular case because concatenation construct pair of result (hence multiple concatenation would produce nested pairs) whereas we want to present its result as a flattened list of arguments suitable for immediate use by a function.

This approach also complicates the usage of Scala parsers because the algebra can only contain a subset of the Scala language can be represented in LMS. Taking the example of matrix multiplication, the user would like not only to solve the dynamic problem on GPU, but also to leverage the graphic card to compute matrices products. Since associated resources description cannot be represented by LMS nodes, the user would need to extend LMS with ad-hoc DSL for every particular function that is not representable in LMS.

\item A second possibility would be to analyze the (user) Scala program and produce another Scala program (simplified) in the restricted subset that is compatible with LMS, hereby providing some ad-hoc transformations either during the conversion (user to simplified), or at a later stage during the optional optimization phase of LMS. This approach seems quite complicated and error prone as it introduces multiple sources of possible errors: during the ad-hoc conversion (user to simplified), during the optimization phase, and during the conversion to C code. Additionally, since dynamic programming require special scheduling to respect the dependencies between intermediate solutions, we would need at a later stage to combine ad-hoc scheduling code with the result of the LMS code generation. 
\item Since ADP normalizes the structure of dynamic programming, the concerns between grammar and algebra are well separated. Since the grammar has a limited number of parsers, it is possible to provide an ad-hoc implementation for each of them thereby allowing better integration. Given that the algebra can contain arbitrary used-defined functions, we avoid duplicating the efforts and rely on LMS to generate the corresponding C code.
\ule
For all the previous reasons it is more simple to provide an ad-hoc code generation for the grammar, whereas the algebra can be generated by LMS, hereby avoiding to duplicate code generation efforts while at the same time allowing us to integrate tightly the grammar parsers into the surrounding scheduling C code.

From an engineering perspective although the features offered by LMS seem very desirable, we should refrain from adding unnecessary levels of complexity\footnote{KISS engineering principle: \url{http://en.wikipedia.org/wiki/KISS_principle}}. Also given the short time frame allowed to this project (with respect to the requirements), I considered that it was preferable to spend time in making progress rather than improving an existing project (but I would reconsider my position given different deadlines). Also during our attempts to move all the parsers on LMS, we also investigated in the Scala 2.10 macros\footnote{(Macros operate directly on the internal representation of the Scala compiler, after the type analysis and and before the code generation phase.} \cite{scala_macros}, which represent a possible alternative to LMS for generating simple functions for a costing algebra.

>>>>

Because dependencies are along line, column or intermediate elements (\S\ref{problem_examples}),  parallel progress can be made along the diagonal of the matrices, a naive row/column addressing strategy would result in no coalesced accesses. Hence we address them by diagonal. Let $M_W$ the width of the matrix and $M_H$ its height. Coalesced addressing is most easy in the parallelogram matrix, as we could pretend that the parallelogram is simply a tilted rectangle  (with diagonal elements being contiguous in memory) as follows:
\[(i,j) \to i + (j-i)*M_H\]

Noticing that the triangle matrix shape is just half the parallelogram, we could use the same addressing scheme, but we would use twice more memory than necessary. Hence we need a tailored formula: the size of the triangle embedded in a square of side $k$ (incl. diagonal) is $\frac{k\cdot(k+1)}{2}$; knowing that fact, the index of the element $(i,j)$ of the triangle (starting with the main diagonal when $i=j$), we obtain the following formula (with $M_H=M_W=n$):
\[(i,j) \to M-T+i \text{ with }\left\{\begin{array}{rcll}
	M&=&\dfrac{(n\cdot (n+1)}{2} \vspace{.15cm} & \textit{total }\bigtriangleup\\
	d&=&n+1+i-j \vspace{.15cm} & \textit{diagonal of }(i,j)\\
	T &=&\dfrac{(d\cdot (d+1)}{2} & \bigtriangleup \textit{ of current diagonal}
\end{array}\right.\]

Finally, for the rectangular matrix, since the parallelogram indexing looks efficient, we want to reuse the same idea. However, embedding the rectangle within a parallelogram would have a very large overhead ($(M_H)^2$, by adding a triangle on each side of the rectangle). The solution consists of breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, thereby dramatically reducing the size of the additional triangles. Finally, the stripes can be stitched together to form a single parallelogram continuing along the next stripe. Noticing that beyond a certain number of lines, coalescing access does not improve latency as memory is anyway not stored contiguously, we can make $B_H$ constant .

@@

% ------------------------------------------------------------------------------------------------
\subsection{LMS integration} \label{lms_use}
At first glance, LMS seems the ideal candidate to transform Scala code into its C-like equivalent. However, the concern in this project is that the GPU code sensibly differs from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment. We want to discuss how LMS could best be used in our project:

LMS only supports a subset of Scala. Embedding both algebrae and grammars into the LMS framework would reduce the expressivity of algebrae. Taking the example of matrix multiplication, the user would like to not only solve the dynamic problem on GPU, but also leverage the graphic card, the disk or the network to compute the matrices products. Since these resources are outside for the LMS world, the user needs to write an ad-hoc DSL for every particular function. In order to avoid this constraint, we need to decouple the grammar and algebra generation.

For the generation of algebra functions, since CUDA types are restricted, they are representable in LMS. Hence we can leverage LMS for CUDA-compatible algebrae whereas regular Scala can be used for complex algebrae being executed on CPU exclusively.

It remains to generate the C code corresponding to the grammar; we argue in favor of a specialized conversion rather than generation through LMS for the following reasons:\ul
\item We have only 6 parser classes to convert (tabulations, terminals, aggregations, filters, alternatives, concatenations), the rest of the user program is part of the algebra. These parsers are combined with very little modifications to create the grammar.
\item The behavior of the CUDA parsers sensibly differs from that available in Scala (\S\ref{normalization}). This would imply writing optimization phases in LMS to convert them appropriately.
\item LMS operations on collections (lists, hash maps, arrays, ...) do not require special scheduling as all elements are treated independently; nevertheless dynamic programming introduces dependencies between elements, thus requiring particular scheduling interleaved with the computation. Enriching LMS with primitives to solve these constraints might be very complex if we want to generalize. However, we could leverage the knowledge that the pattern belongs to dynamic programming, hereby reducing the complexity of the work and possibly being more efficient. Additionally, a specific memory pattern is required to increase efficiency (\S\ref{mem_layout}), we also need to reuse this knowledge to LMS, so that it can generate optimal code, so we might end up writing a specialized code generator.
\item Finally, LMS is a rather complicated framework, with sparse documentation. Although many features are available (support for tuples, objects, matrices, ...), it is not easy to figure out where to find them (because they may reside on different branches or are not tested), hence the time taken to improve the framework to suit our needs far outweighs the time to build a tailored solution, which is the choice we are somehow forced to take, given the time constraints of this project.
\ule
Since we need to operate a lot of transformation, inside LMS or in specialized generation, the choice of specialized code generation was guided by the engineering principle of steering away from unnecessary complexity\footnote{KISS engineering principle: \url{http://en.wikipedia.org/wiki/KISS_principle}}, hence reducing the number of potential sources of errors. During our experiments with LMS, we also investigated in the Scala 2.10 macros\footnote{(Macros operate directly on the internal representation of the Scala compiler, after the type analysis and and before the code generation phase.} \cite{scala_macros}, which represent a possible alternative to LMS for generating very simple functions for a costing algebra.

### 2013-01-18 docs/report/introduction.tex
Nits.

% ------------------------------------------------------------------------------------------------
\newpage
\setcounter{tocdepth}{2} \tableofcontents
\newpage
\section{Introduction} \label{intro}
Dynamic programming (DP) is an algorithmic technique to solve optimization problems. For example, we might want to multiply a chain of matrices\footnote{Such that matrices have appropriate dimensions to be multiplied with each other.} efficiently. The order in which matrices are combined changes the number of required scalar multiplications, we would want to find an optimal order that minimizes this number. Notice that if knew how to split of the chain into two subparts, we could recursively find an optimal order in these two parts and recombine them. Such combinatorial problems verify the Bellman's principle\cite{bellman_principle}: \textit{<<optimal solutions depends on optimal solutions of sub-problems>>}\footnote{\url{http://en.wikipedia.org/wiki/Bellman\_equation\#Bellman.27s\_Principle\_of\_Optimality}}. Because we don't know a priori where we could split the chain, we would need to explore an exponential number of possibilities. Using the Bellman's principle, we can memoize intermediate optimal solutions to save redundant computations, hereby reducing the problem to a polynomial complexity.

@@

% ------------------------------------------------------------------------------------------------
\section{Background} \label{background}
\subsection{Graphic cards}
Modern graphic cards\footnote{We cover here interesting features of the CUDA devices and programming paradigm; however, the same concept should be applicable to graphic cards from other vendors.} are powered by massively parallel processor: they can typically run hundreds or thousands of cores, each able to schedule multiple threads. The threads are usually grouped in warps that are scheduled synchronously. This means that if there is a divergence in the execution path, both alternative are executed sequentially, hereby stalling other warp's threads. Threads are logically grouped in blocks by the programmer whereas warps correspond to a physical constraint. In a deliberate design decision to simplify the hardware, there exist no global synchronization.

>>>>

% ------------------------------------------------------------------------------------------------
\newpage
\setcounter{tocdepth}{2} \tableofcontents
\newpage
\section{Introduction} \label{intro}
Dynamic programming (DP) is an algorithmic technique to solve optimization problems. For example, we might want to multiply a chain of matrices\footnote{Such that matrices have appropriate dimensions to be multiplied with each other.} efficiently. The order in which matrices are combined changes the number of required scalar multiplications, we would want to find an optimal order that minimizes this number. Notice that if knew how to split of the chain into two subparts, we could recursively find an optimal order in these two parts and recombine them. Such combinatorial problems verify the Bellman's principle\cite{bellman_principle}: \textit{<<optimal solutions depends on optimal solutions of sub-problems>>}\footnote{\url{http://en.wikipedia.org/wiki/Bellman\_equation\#Bellman.27s\_Principle\_of\_Optimality}}. Because we don't know a priori where we could split the chain, we would need to explore an exponential number of possibilities. Using the Bellman's principle, we can memoize intermediate optimal solutions to save redundant computations, thereby reducing the problem to a polynomial complexity.

@@

% ------------------------------------------------------------------------------------------------
\section{Background} \label{background}
\subsection{Graphic cards}
Modern graphic cards\footnote{We cover here interesting features of the CUDA devices and programming paradigm; however, the same concept should be applicable to graphic cards from other vendors.} are powered by massively parallel processors: they can typically run hundreds or thousands of cores, each able to schedule multiple threads. The threads are usually grouped in warps that are scheduled synchronously. This means that if there is a divergence in the execution path, both alternatives are executed sequentially, thereby stalling other warp's threads. Threads are logically grouped in blocks by the programmer whereas warps correspond to a physical constraint. In a deliberate design decision to simplify the hardware, there exist no global synchronization.

### 2013-01-18 docs/report/problems.tex
Nits.

% ----------------------------------------------
\newpage
\subsubsection{Zuker RNA folding} \label{zuker} \ol
\item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy (which is based on actual measurements, hereby much more complicated than Nussinov \S\ref{nussinov}).
\item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
	\item Cost matrices:\ul
		\item $\Sigma(W)=\Sigma(V)=\{0..z\}$ with $z \le n \cdot b+c$
		\item $\Sigma(F)=\{0..y\}$ with $y\le \min(F_0, z\cdot n)$
		\ule
	\item Backtrack matrices: \ul
		\item $\Sigma(BW)=\{stop, L,R,V,k\}$
		\item $\Sigma(BV)=\{stop, hairpin, stack, (i,j) , k\}$ with $0\le i,j,k < n$ % \\ $HL$=HairpinLoop, $IL$=InteriorLoop, $(i,j)$=MultiLoop
		\item $\Sigma(BF)=\{stop, P, k\}$ with $0\le k < n$
		\ule
	\ule
\item Initialization:\ul
	\item Cost matrices: $W_{(i,i)}=V_{(i,i)}=0, F_{(0)}=$ energy of the unfolded RNA.
	\item Backtrack matrices: $BW_{(i,i)}=BV_{(i,i)}=BF_{(0)}=stop$.
	\ule
\item Recurrence:
\[\begin{array}{rcl}
W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	W_{(i+1,j)}+b & L\\
	W_{(i,j-1)}+b & R\\
	V_{(i,j)}+\delta(S_i,S_j) & V \\
	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} &k
\end{array}\right\} = BW_{(i,j)}\\
\\
V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
	eh(i,j)+b \qquad\qquad\, \text{otherwise} & hairpin\\
	V_{(i+1,j-1)}+es(i,j) & stack \\
	VBI_{(i,j)} & (i',j') \\
	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
\end{array}\right\} = BV_{(i,j)}\\
\\
VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}\\
\\
F_{(j)}&=&\min\left\{\begin{array}{l|l}
	F_{(j-1)} & P \\
	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
\end{array}\right\} = BF_{(j)}
\end{array}\]

>>>>

% ----------------------------------------------
\newpage
\subsubsection{Zuker RNA folding} \label{zuker} \ol
\item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy (which is based on actual measurements, hence much more complicated than Nussinov \S\ref{nussinov}).
\item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
	\item Cost matrices:\ul
		\item $\Sigma(W)=\Sigma(V)=\{0..z\}$ with $z \le n \cdot b+c$
		\item $\Sigma(F)=\{0..y\}$ with $y\le \min(F_0, z\cdot n)$
		\ule
	\item Backtrack matrices: \ul
		\item $\Sigma(BW)=\{stop, L,R,V,k\}$
		\item $\Sigma(BV)=\{stop, hairpin, stack, (i,j) , k\}$ with $0\le i,j,k < n$ % \\ $HL$=HairpinLoop, $IL$=InteriorLoop, $(i,j)$=MultiLoop
		\item $\Sigma(BF)=\{stop, P, k\}$ with $0\le k < n$
		\ule
	\ule
\item Initialization:\ul
	\item Cost matrices: $W_{(i,i)}=V_{(i,i)}=0, F_{(0)}=$ energy of the unfolded RNA.
	\item Backtrack matrices: $BW_{(i,i)}=BV_{(i,i)}=BF_{(0)}=stop$.
	\ule
\item Recurrence:
\[\begin{array}{rcl}
W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	W_{(i+1,j)}+b & L\\
	W_{(i,j-1)}+b & R\\
	V_{(i,j)}+\delta(S_i,S_j) & V \\
	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} &k
\end{array}\right\} = BW_{(i,j)}\\
\\
V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
	eh(i,j)+b \qquad\qquad\, \text{otherwise} & hairpin\\
	V_{(i+1,j-1)}+es(i,j) & stack \\
	VBI_{(i,j)} & (i',j') \\
	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
\end{array}\right\} = BV_{(i,j)}\\
\\
VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}\\
\\
F_{(j)}&=&\min\left\{\begin{array}{l|l}
	F_{(j-1)} & P \\
	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
\end{array}\right\} = BF_{(j)}
\end{array}\]
